Quantitative processing of broadband data as implemented in a scientific splitbeam echosounder

### 2022-03-18 Paper/Manuscript.tex
Small fixes to paper

@@

Active acoustics is an efficient method for remote sensing of marine ecosystems. These methods have been efficiently used to cover a wide range of spatial and temporal scales from the aquatic ecosystem, ranging from tens of km (Makris) to small scale behaviour patterns (refs). These methods are also a crucial component of fisheries resource management. Several different configurations of acoustic instruments are available, these include multibeam sonars, synthetic aperature sonar, omnidirectional sonars, acoustic imaging sonars as well as echosounders (refs).

Single- beam echosounders have extensively been used in fisheries management and ecosystem monitoring. Methodology for using these systems are well developed and in particular the ability for accurate calibration of the system has made these specially useful. Single-beam echosounders have provided important information for fisheries resource management (Simmonds and MacLennan, 2005) as well as insight in ecological processes for the aquatic ecosystem (Godø 2014).

Conventional echosounders generate an acoustic pulse with a narrow bandwidth (several kHz at most), and the simultaneous operation at widely spaced frequencies (such as 18, 38, 70, 120, 200, and 333 kHz). The method has been used to  study predator-prey interactions (Benoit-Bird and Au, 2003; Friedlaender et al., 2009; Veit et al., 1993), interactions between marine organisms and the physical environment (Crawford et al., 2012; Freeman et al., 2004; Georgiadis et al., 2009; Ludsin et al., 2009; Godø...), abundance and distribution of prey and predators (Thompson et al., 1991) and non-fish species or group of species (Brierley et al., 2001), and the individual or group behavior of marine organisms (Axelsen et al., 2000; Benoit-Bird et al., 2001; Kaartvedt et al., 2009).

Recent echosounders produce pulses with a wide and continuous frequency range (broadband pulses), as opposed to conventional single frequency systems, i.e. a continuous frequency coverage over a wide frequency band. provide significantly better along-beam resolution and a higher signal to noise ratio than narrowband pulses (Chu and Stanton, 1998; Ehrenberg and Torkelson, 2000). These benefits can lead to improved backscatter categorization (R. Korneliussen, 2018; R. J. Korneliussen et al., 2016; Martin Traykovski et al., 1998; Stanton et al., 2012) and hence more accurate distribution and abundance estimates, as well as enhanced information on ecosystem structure and functioning 
(skriv om: e.g. Lavery et al., 2017; Basset et al., 2017; 2020; Skaret et al., 2020; Kubilius et al., 2020; Benoit-Bird and Waluk, 2020; Hasegawa et al., 2021; Gugele et al., 2021; Cotter et al., 2021; Blaunet et al., 2021; Agersted et al., 2021; Khodabandeloo et al., 2021a;2021b).

@@

To tap the full potential of broadband echosounder data, further methods for processing such data must be developed. Examples include adapting pulses to improve target separation as well as further parameters from the individual targets. Different send pulses can be envisioned as well as various methods for acoustic target classification. The objective of this paper is to present a systematic and comprehensive explanation of how to process calibrated broadband echosounder data. By making associated code available, we provide a starting point for further developing signal processing methods for applying broadband acoustic to aquatic ecology.

@@

In the example a linear sweep enveloped by a Hann function is implemented (Fig.~\ref{fi:ytx}). This is found in the class \verb|Core.EK80Calculation.generateIdealWindowedSendPulse|, where the parameters are the initial frequency, the final frequency, the pulse duration, the sampling rate, and the proportion of the signal that is tapered in each end, respectively. A tapering of 0 and .5 indicates no tapering and tapering across the whole signal, respectively.
\begin{figure}
\includegraphics[width=16cm]{Fig_ytx}
\caption{\label{fi:ytx} Ideal and enveloped send pulse between 34kHz to 45kHz with a slope of 0.057, as implemented in the EK80. The orange curve illustrate the effect of setting the slope to 0.5.}
\end{figure}

@@

\begin{figure}
\includegraphics[width=16cm]{Fig_fir}
\caption{\label{fi:fir} The frequency (power) response of the filters in from our test set.}
\end{figure}

The original sample data $\yrxorg(\samplesymt,\channelsym)$ are not available in the EK80 data files. Instead,the resulting filtered and decimated complex samples from each transducer channel $\yrx(\samplesymt,\channelsym)$ are stored in the data files. The data are recorded together with additional data such as from position and motion sensors and system configuration data in raw data files for display and analysis by processing software. The filter parameters are in this paper used to generate the transmit pulse for the matched filter, see below.

@@

\begin{figure}
\includegraphics{Fig_ACF}
\caption{The aoutocorrekation function $\ptxauto$ from the example. The corresponding $\teff=0.1ms$ and $\tnom=2ms$\label{fi:ACF}.}
\end{figure}

@@

Forward/aft and port/starboard phase angles of target echoes are estimated by combining the transducer half signals thus: 
%
\begin{eqnarray}
\label{eq:phase1}
y_\along(\samplesymt) & = & y_{\textrm{pc,fore}}(\samplesymt) y_{\textrm{pc,aft}}^*(\samplesymt), \\
y_\athw(\samplesymt) & = & y_{\textrm{pc,star}}(\samplesymt) y_{\textrm{pc,port}}^*(\samplesymt),
\end{eqnarray}
%
where $y_\along(\samplesymt)$ is the electrical angle along the minor axis of the transducer (positive in the forward direction when ship-mounted) and $y_\athw(\samplesymt)$ the electrical angle along the major axis of the transducer (positive to starboard when ship-mounted), where complex signals are represented in the form $e^{j 2\pi \freqsym \timesym}$, where $j = \sqrt{-1}$. The physical echo arrival angles ($\along$ and $\athw$) are then given by:
%
\begin{eqnarray}
\label{eq:phase2}
\along(\samplesymt) & = & \arcsin\left( \frac{\atan\left( \Im(y_\along(\samplesymt)), \Re(y_\along(\samplesymt) \right)}{\anglefalong}\right) \\
\athw(\samplesymt) & = & \arcsin\left( \frac{\atan\left( \Im(y_\athw(\samplesymt)), \Re(y_\athw(\samplesymt) \right)}{\anglefathw}\right).
\end{eqnarray}
%
where $\anglefalong$ and $\anglefathw$ are constants that convert from phase angles to physical echo arrival angles and are derived from the transducer geometry and $\fc$ the centre frequency of the chirp pulse \citep{ehrenberg1979}. The inverse sine is indicated by $\arcsin$, the four quadrant inverse tangent which returns values in the interval $[-\pi, \pi]$ inclusive is indicated by $\atan$, the real part of a complex number by $\Re$ and the imaginary part by $\Im$. As a mnemonic, the horizontal line in the symbol used for the forward/aft direction, $\along$, represents the pivot axis for the alongship angles and the near-vertical line in the $\athw$ symbol indicates the pivot axis for port/starboard angles.
%
\begin{figure}
\includegraphics[width=16cm]{Fig_theta_phi}
\caption{\label{fi:theta_phi}Physical angles.}
\end{figure}

@@

To illustrate each step of the process to obtain $TS(f)$ we use a single ping from a data file where a single metallic calibration sphere is suspended approximately 6 m below a 120 kHz transducer (fig.~\ref{Fig_TS_echogram}).

\begin{figure}
\includegraphics[width=16cm]{Fig_TS_echogram}
\caption{\label{Fig_TS_echogram} Sv as a function of frequency ($m$) and range ($n$) for a echogram raw data file. The red vertical line indicates ping to be processed in this section for TS(f). Single target calibration sphere is seen at approximately 6 m range and the seafloor at approximately 12 m.}
\end{figure}

@@

The discrete Fourier transforms of the target signal, $\ypctargetf(\samplesymf)$, and the reduced auto correlation signal, $\ymfautoredf(\samplesymf)$, are given by:
\begin{eqnarray}
\label{eq:DFT_Target_Auto}
\ypctargetf(\samplesymf) & = & \dft_\ndft(\ypctarget(\samplesymt)),\\
\ymfautoredf(\samplesymf) & = & \dft_\ndft(\ymfautored(\samplesymt)),
\end{eqnarray}
where $\dft$ indicates the Fourier transform of length $\ndft$ and $\samplesymf$ the sample index in the frequency domain.
The nomalized discrete Fourier transform of the target signal, $\ypctargetnormf(\samplesymf)$, (\ref{fi:SED}) is then calculated by: 
%
\begin{equation}
\label{eq:DFT_Target_Auto_Norm}
\ypctargetnormf(\samplesymf) = \frac{\ypctargetf(\samplesymf)} {\ymfautoredf(\samplesymf)}.
\end{equation}
%
The corresponding frequencies f\_m\_t (?) are calculate as follows... See the function EK80CalculationPaper.calcDFTforTS. Nils Olav did not quite get this so please enlighten me...

@@

A frequency modulated pulse scattered by a metallic sphere will exhibit frequencies at which very little energy is returned due to destructive interference \citep{stanton2008}. This is visible in the $\ts$ (Fig. \ref{fi:SED}) and agrees well with theoretical estimates of the backscatter from spheres \citep{maclennan1981}. The amplitude of the backscatter signal also clearly shows these nulls (Fig. \ref{fi:ts_example}c), which are readily visible here due to the use of a linear chirp where time through the pulse corresponds to specific frequencies.

Comment from Nils Olav: Using $m$ as an argument to a function is not ideal. If we use the function argument perhaps using $f_m$ would be more logical, i.e. $\lambda(f_m)$. Alternatively using $lambda_m$ is possible. The latter is perhaps more in line with the code since the parameter is an array windexed with $m$ and not a function. $Psi(f)$, on the the other hand is derived from a function, see below.

@@

As a metallic sphere is a rather simple and ideal scatterer, we present $\sv$ from a school of non-swimbladdered fish (\textit{tobis}) (fig.~\ref{Fig_Sv_echogram}) in the processes to obtain $Sv((f)$.

@@

TODO: Explain edge cases.

@@

%
TODO: How to index samples along range? In the code \_n is used to indicate the range component of the FFt's. Needs attention.

@@

By selecting a set of centre samples $n$, $Sv$ values can be presented as a function of range (n) and frequency (m) for each ping. The range for centre samples n could be chosen to half the window length or any other grid that the user prefer the data presented to be in. This can be useful when combining the Sv(f) across a range of transducers. In our example we have simply chosen n to be on the same order of the range resolution (fig.~\ref{Fig_Sv_m_n}).

@@

For echo integration purposes it is common to integrate the Sv over a range, in our example we integrate the Sv from 15 to 34 m covering a school of non-swimbladdered fish (fig). We note that we normally require averaging over several individual pings to obtain an unbiased estimate of Sv for echo integration purposes. Here we present the sv(f) over the layer for our example ping for illustrative purposes (fig.~\ref{Fig_Sv_avg}). Even though this is for a single ping, we still observe a positive slope of the frequency response indicative of non-swimbladdered fish.

\begin{figure}
\includegraphics[width=16cm]{Fig_Sv_avg}
\caption{\label{Fig_Sv_avg} Sv as a function of frequency ($m$) averaged over a range interval covering a fish school }
\end{figure}

>>>>

\usepackage{color,soul}

@@

Active acoustics is an efficient method for remote sensing of marine ecosystems. These methods have been efficiently used to cover a wide range of spatial and temporal scales from the aquatic ecosystem, ranging from tens of km (Duane et al., 2021) to small scale behaviour patterns (Handegard et al., 2003). These methods are also a crucial component of fisheries resource management. Several different configurations of acoustic instruments are available, these include multibeam sonars, synthetic aperture sonar, omnidirectional sonars, acoustic imaging sonars as well as single- and multibeam echosounders (\hl{refs}).

Single-beam echosounders have extensively been used in fisheries management and ecosystem monitoring. Methodology for using these systems are well developed and in particular the ability for accurate calibration of the system has made these specially useful. Single-beam echosounders have provided important information for fisheries resource management (Simmonds and MacLennan, 2005) as well as insight in ecological processes for the aquatic ecosystem (Godø, 2014).

Conventional echosounders generate an acoustic pulse with a narrow bandwidth (several kHz at most), and the simultaneous operation at widely spaced frequencies (such as 18, 38, 70, 120, 200, and 333 kHz). The method has been used to  study predator-prey interactions (Benoit-Bird and Au, 2003; Friedlaender et al., 2009; Veit et al., 1993), interactions between marine organisms and the physical environment (Crawford et al., 2012; Freeman et al., 2004; Georgiadis et al., 2009; Ludsin et al., 2009; Godø, 2014), abundance and distribution of prey and predators (Thompson et al., 1991) and non-fish species or group of species (Brierley et al., 2001), and the individual or group behavior of marine organisms (Axelsen et al., 2000; Benoit-Bird et al., 2001; Kaartvedt et al., 2009).

Recent echosounders produce pulses with a wide and continuous frequency range (broadband pulses), as opposed to conventional single frequency systems, i.e. a continuous frequency coverage over a wide frequency band. This provides significantly better along-beam resolution and a higher signal to noise ratio than narrowband pulses (Chu and Stanton, 1998; Ehrenberg and Torkelson, 2000). These benefits can lead to improved backscatter categorization (R. Korneliussen, 2018; R. J. Korneliussen et al., 2016; Martin Traykovski et al., 1998; Stanton et al., 2012) and hence more accurate distribution and abundance estimates, as well as enhanced information on ecosystem structure and functioning. Early work with recent commercial systems indicate the potential improvement in detection near boundaries (Lavery et al., 2017) and remote sizing of fish (Kubilius et al., 2020;), capabilities of fine-scale observations of interactions between prey and predators (Skaret et al., 2020), the potential for species discrimination (Basset et al., 2018; Blanluet et al., 2019; Benoit-Bird and Waluk, 2020; Hasegawa et al., 2021; Gugele et al., 2021), and the potential for enhanced information on mesopelagic layers (Cotter et al., 2021; Agersted et al., 2021; Khodabandeloo et al., 2021a;2021b).

@@

To tap the full potential of broadband echosounder data, further methods for processing such data must be developed (Lavery et al., 2017; Basset et al., 2018). Examples include adapting pulses to improve target separation as well as further parameters from the individual targets. Different send pulses can be envisioned as well as various methods for acoustic target classification. The objective of this paper is to present a systematic and comprehensive explanation of how to process calibrated broadband echosounder data. By making associated code available, we provide a starting point for further developing signal processing methods for applying broadband acoustic to aquatic ecology.

@@

In the example a linear sweep enveloped by a Hann function is implemented (Fig.~\ref{fi:ytx}). This is found in the class \verb|Calculation.generateIdealWindowedTransmitSignal|, where the parameters are the initial frequency, the final frequency, the pulse duration, the sampling rate, and the proportion of the signal that is tapered in each end, respectively. A tapering of 0 and .5 indicates no tapering and tapering across the whole signal, respectively.
\begin{figure}
\includegraphics[width=16cm]{Fig_ytx}
\caption{\label{fi:ytx} Ideal and enveloped transmit pulse between 34 kHz to 45 kHz with a slope of 0.057, as implemented in the EK80. The orange curve illustrate the effect of setting the slope to 0.5.}
\end{figure}

@@

\begin{figure}
\includegraphics[width=16cm]{Fig_fir}
\caption{\label{fi:fir} The frequency (power) response of the filters in our test set.}
\end{figure}

The original sample data $\yrxorg(\samplesymt,\channelsym)$ are not available in the EK80 data files. Instead, the resulting filtered and decimated complex samples from each transducer channel $\yrx(\samplesymt,\channelsym)$ are stored in the data files. The data are recorded together with additional data such as from position and motion sensors and system configuration data in raw data files for display and analysis by processing software. The filter parameters are in this paper used to generate the transmit pulse for the matched filter, see below.

@@

\begin{figure}
\includegraphics{Fig_ACF}
\caption{The autocorrelation function $\ptxauto$ from the example. The corresponding $\teff=0.1 ms$ and $\tnom=2 ms$\label{fi:ACF}.}
\end{figure}

@@

Forward/aft and port/starboard phase angles of target echoes are estimated by combining the transducer half signals thus: 
%
\begin{eqnarray}
\label{eq:phase1}
y_\along(\samplesymt) & = & y_{\textrm{pc,fore}}(\samplesymt) y_{\textrm{pc,aft}}^*(\samplesymt), \\
y_\athw(\samplesymt) & = & y_{\textrm{pc,star}}(\samplesymt) y_{\textrm{pc,port}}^*(\samplesymt),
\end{eqnarray}
%
where $y_\along(\samplesymt)$ is the electrical angle along the minor axis of the transducer (positive in the forward direction when ship-mounted) and $y_\athw(\samplesymt)$ the electrical angle along the major axis of the transducer (positive to starboard when ship-mounted), where complex signals are represented in the form $e^{j 2\pi \freqsym \timesym}$, where $j = \sqrt{-1}$. The physical echo arrival angles ($\along$ and $\athw$) are then given by:
%
\begin{eqnarray}
\label{eq:phase2}
\along(\samplesymt) & = & \arcsin\left( \frac{\atan\left( \Im(y_\along(\samplesymt)), \Re(y_\along(\samplesymt) \right)}{\anglefalong}\right) \\
\athw(\samplesymt) & = & \arcsin\left( \frac{\atan\left( \Im(y_\athw(\samplesymt)), \Re(y_\athw(\samplesymt) \right)}{\anglefathw}\right).
\end{eqnarray}
%
where $\anglefalong$ and $\anglefathw$ are constants that convert from phase angles to physical echo arrival angles and are derived from the transducer geometry and $\fc$ the centre frequency of the chirp pulse \citep{ehrenberg1979}. The inverse sine is indicated by $\arcsin$, the four quadrant inverse tangent which returns values in the interval $[-\pi, \pi]$ inclusive is indicated by $\atan$, the real part of a complex number by $\Re$ and the imaginary part by $\Im$. As a mnemonic, the horizontal line in the symbol used for the forward/aft direction, $\along$, represents the pivot axis for the alongship angles and the near-vertical line in the $\athw$ symbol indicates the pivot axis for port/starboard angles.
%
\begin{figure}
\includegraphics[width=16cm]{Fig_theta_phi}
\caption{\label{fi:theta_phi}Physical angles $\along$ and $\athw$.}
\end{figure}

@@

To illustrate each step of the process to obtain $TS(m)$ we use a single ping from a data file where a single metallic calibration sphere with 35 mm diameter (WC35) is suspended approximately 6 m below a 120 kHz transducer (fig.~\ref{Fig_TS_echogram}).

\begin{figure}
\includegraphics[width=16cm]{Fig_TS_echogram}
\caption{\label{Fig_TS_echogram} Sv as a function of frequency ($m$) and range ($n$) for a echogram raw data file. The red vertical line indicates ping to be processed in this section for TS(m). Single target calibration sphere is seen at approximately 6 m range and the seafloor at approximately 12 m.}
\end{figure}

@@

The discrete Fourier transforms of the target signal, $\ypctargetf(\samplesymf)$, and the reduced auto correlation signal, $\ymfautoredf(\samplesymf)$, are given by:
\begin{eqnarray}
\label{eq:DFT_Target_Auto}
\ypctargetf(\samplesymf) & = & \dft_\ndft(\ypctarget(\samplesymt)),\\
\ymfautoredf(\samplesymf) & = & \dft_\ndft(\ymfautored(\samplesymt)),
\end{eqnarray}
where $\dft$ indicates the Fourier transform of length $\ndft$ and $\samplesymf$ the sample index in the frequency domain.
The nomalized discrete Fourier transform of the target signal, $\ypctargetnormf(\samplesymf)$, (\ref{fi:SED}) is then calculated by: 
%
\begin{equation}
\label{eq:DFT_Target_Auto_Norm}
\ypctargetnormf(\samplesymf) = \frac{\ypctargetf(\samplesymf)} {\ymfautoredf(\samplesymf)}.
\end{equation}
%
\hl{The corresponding frequencies f\_m\_t (?) are calculate as follows... See the function EK80CalculationPaper.calcDFTforTS. Nils Olav did not quite get this so please enlighten me...}

@@

A frequency modulated pulse scattered by a metallic sphere will exhibit frequencies at which very little energy is returned due to destructive interference \citep{stanton2008}. This is visible in the $\ts$ (Fig. \ref{fi:SED}) and agrees well with theoretical estimates of the backscatter from spheres \citep{maclennan1981}. The amplitude of the backscatter signal also clearly shows these nulls (Fig. \ref{fi:SED}), which are readily visible here due to the use of a linear chirp where time through the pulse corresponds to specific frequencies.

\hl{Comment from Nils Olav: Using $m$ as an argument to a function is not ideal. If we use the function argument perhaps using $f_m$ would be more logical, i.e. $\lambda(f_m)$. Alternatively using $lambda_m$ is possible. The latter is perhaps more in line with the code since the parameter is an array windexed with $m$ and not a function. $Psi(f)$, on the the other hand is derived from a function, see below.}

@@

As a metallic sphere is a rather simple and ideal scatterer, we use $\sv$ from a school of non-swimbladdered fish (\textit{tobis}) (fig.~\ref{Fig_Sv_echogram}) to illustrate the processes to obtain $Sv(m)$.

@@

\hl{TODO: Explain edge cases.}

@@

%
\hl{TODO: How to index samples along range? In the code \_n is used to indicate the range component of the FFt's. Needs attention.}

@@

By selecting a set of centre samples $n$, $Sv$ values can be presented as a function of range ($n$) and frequency ($m$) for each ping. The range for centre samples $n$ could be chosen to half the window length or any other grid that the user prefer the data presented to be in. This can be useful when combining the $Sv(m)$ across a range of transducers. In our example we have simply chosen n to be on the same order of the range resolution (fig.~\ref{Fig_Sv_m_n}).

@@

For acoustic abundance estimation and classification purposes it is common to integrate the Sv over a range, in our example we integrate $Sv$ from 15 to 34 m covering a school of non-swimbladdered fish (fig.~\ref{Fig_Sv_echogram}). We note that we normally require averaging over several pings to obtain an unbiased estimate of $Sv$ for echo integration purposes. Here we present the $Sv(m)$ over the layer for our example ping for illustrative purposes (fig.~\ref{Fig_Sv_avg}). Even though this is for a single ping, we still observe a positive slope of the frequency response indicative of non-swimbladdered fish.

\begin{figure}
\includegraphics[width=16cm]{Fig_Sv_avg}
\caption{\label{Fig_Sv_avg} $Sv$ as a function of frequency ($m$) averaged over a range interval covering a fish school }
\end{figure}
